A Multi-Scale Analysis of 27,000 Urban Street Networks

### 2017-12-14 draft/v1/latex/paper.tex
revision

\documentclass{article}

\usepackage{graphicx}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{pdflscape}
\usepackage{setspace}
\usepackage[USenglish]{babel}
\usepackage{csquotes}
\usepackage[numbers]{natbib}
\usepackage[affil-it]{authblk}

\doublespacing
\bibliographystyle{chicago}

\begin{document}
\title{A Multi-Scale Analysis of 27,000 Urban Street Networks\footnote{This is a journal article manuscript currently under peer review.}}
\author{Geoff Boeing\footnote{Email address: gboeing@berkeley.edu; corresponding author.}}
\affil{Department of City and Regional Planning\\University of California, Berkeley}
\date{\today}
\maketitle

@@

OpenStreetMap offers an underexplored source of worldwide geospatial data useful to urban researchers. This study uses the OSMnx software to automatically download and analyze over 27,000 street networks from OpenStreetMap at metropolitan, municipal, and neighborhood scales---namely, every US city and town ($N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$). It presents wide-ranging empirical findings on US urban form and street network characteristics, emphasizing measures relevant to urban design and morphology such as structure, connectedness, density, centrality, and resilience. In the past, street network data acquisition has been challenging and ad hoc, but this study illustrates the use of OSMnx and OpenStreetMap to conduct street network analysis with extremely large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. These street networks and their measures are now shared in a public repository for other researchers to use.

@@

On May 20, 1862, Abraham Lincoln signed the Homestead Act into law, making land across the United States' Midwest and Great Plains available for free to applicants \cite{porterfield_homestead_2005}. Under its auspices over the next 70 years, the federal government distributed 10\% of the entire US landmass to private owners in the form of 1.6 million homesteads \cite{lee_kansas_1979, sherraden_inclusion_2005}. New towns with gridiron street networks sprang up rapidly across the Great Plains and Midwest, due to both the prevailing urban design paradigm of the day and the standardized rectilinear town plats used repeatedly to lay out instant new cities \cite{southworth_streets_1997}. Through path dependence, the spatial signatures of these land use laws, design paradigms, and planning instruments can still be seen today in these cities' urban forms and street networks.

A cross-sectional analysis of American urban form can reveal these artifacts and histories through street networks at metropolitan, municipal, and neighborhood scales. Network analysis is a natural approach to the study of cities as complex systems \cite{masucci_random_2009}. The empirical literature on street networks is growing ever richer, but suffers from some limitations---discussed in detail in \cite{boeing_osmnx:_2017} and summarized here. First, sample sizes tend to be fairly small due to data availability, gathering, and processing constraints: most studies in this literature that conduct topological or metric analysis tend to have sample sizes ranging around 10 to 50 networks \cite{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, strano_urban_2013}, which may limit the generalizability and interpretability of findings. Second, reproducibility has been difficult when the dozens of decisions that go into analysis---such as spatial extents, topological simplification and correction, definitions of nodes and edges, etc.---are ad hoc or only partly reported \citep[e.g.,][]{porta_network_2006, strano_urban_2013}. Third, and related to the first two, studies frequently oversimplify to planar or undirected primal graphs for tractability \cite[e.g.,][]{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, masucci_random_2009}, or use dual graphs despite the loss of geographic and metric information \cite{batty_network_2005, jiang_integration_2002, ratti_space_2004, crucitti_centrality_2006, crucitti_centrality_2006-1}.

This study addresses these limitations by conducting an urban morphological analysis of street networks at multiple scales, with large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. In particular, it examines urban street networks---represented as primal, nonplanar, weighted multidigraphs with possible self-loops---focusing on structure, connectedness, centrality, and resilience. This study analyzes 27,009 urban street networks at multiple overlapping scales across the US. Namely, it examines the street networks of every incorporated city and town (population size $N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$) in the US. To do so, it uses OSMnx---a new street network research toolkit \cite{boeing_osmnx:_2017}---to download, construct, and analyze these street networks at metropolitan, municipal, and neighborhood scales. These street networks and their measures data have been shared in a public repository for other researchers to use at [link redacted].

The purpose of this paper is threefold. First, it describes and demonstrates a new methodology for easily and consistently acquiring, constructing, and analyzing large samples of street networks as nonplanar directed graphs. Second, it presents empirical findings of descriptive urban morphology for the street networks of every US city, urbanized area, and Zillow neighborhood. Third, it investigates with large sample sizes some claims previously made in the research literature with small sample sizes. This paper is organized as follows. In the next section, it discusses the data sources, tools, and methods used to collect, construct, and analyze these street networks. Next, it presents findings of the analyses at metropolitan, municipal, and neighborhood scales. Finally, it concludes with a discussion of these findings and their implications for street network analysis, urban morphology, and city planning.

@@

A network (also called a \emph{graph} in mathematics) comprises a set of nodes connected to one another by a set of edges. Street networks can be conceptualized as primal, directed, nonplanar graphs. A primal street network represents intersections as nodes and street segments as edges. A directed network has directed edges: that is, edge $uv$ points one-way from node $u$ to node $v$, but there need not exist a reciprocal edge $vu$. A planar network can be represented in two dimensions with its edges intersecting only at nodes \cite[e.g.,][]{viana_simplicity_2013, fischer_spatial_2014}. Most street networks are nonplanar---due to grade-separated expressways, overpasses, bridges, tunnels, etc.---but most quantitative studies of urban street networks represent them as planar \cite[e.g.,][]{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, masucci_random_2009, strano_urban_2013} for tractability because bridges and tunnels are uncommon in some cities. Planar graphs may reasonably model the street networks of old European town centers, but poorly model the street networks of modern autocentric cities like Los Angeles or Tokyo with many grade-separated expressways, bridges, and underpasses.

@@

This study uses OSMnx to download, construct, correct, analyze, and visualize street networks at metropolitan, municipal, and neighborhood scales. OSMnx is a Python-based research tool that easily downloads OpenStreetMap data for any place in the world and constructs it into a spatially-embedded graph-theoretic object for analysis and visualization \cite{boeing_osmnx:_2017}. OpenStreetMap is a collaborative worldwide mapping project that makes its spatial data available via various APIs \cite{corcoran_analysing_2013, jokar_arsanjani_openstreetmap_2015}. These data are of high quality and compare favorably to CIA World Factbook estimates and US Census TIGER/Line data \cite{haklay_how_2010, over_generating_2010, zielstra_comparative_2011, maron_how_2015, wu_improving_2005, frizzelle_importance_2009}. In 2007, OpenStreetMap imported the TIGER/Line roads (2005 vintage) and since then, many community-led corrections and improvements have been made \cite{willis_openstreetmap_2008}. Many of these additions go beyond TIGER/Line's scope, including passageways between buildings, footpaths through parks, bike routes, and detailed feature attributes such as finer-grained street classifiers, speed limits, etc.

To define the study sites and their spatial boundaries, this study uses three sets of geometries. The first defines the metropolitan-scale study sites using the 2016 TIGER/Line shapefile of US census bureau urban areas. Each census-defined urban areas comprises a set of tracts that meet a minimum density threshold \cite{u.s._census_bureau_2010_2010}. We retain only the \emph{urbanized areas} subset of these data (i.e., areas with greater than 50,000 population), discarding the small \emph{urban clusters} subset. The second set of geometries defines our municipal-scale study sites using 51 separate TIGER/Line shapefiles (again, 2016) of US census bureau \emph{places} within all 50 states plus DC. We discard the subset of \emph{census-designated places} (i.e., small unincorporated communities) in these data, while retaining every US city and town. The third set of geometries defines the neighborhood-scale study sites using 42 separate shapefiles from Zillow, a real estate database company. These shapefiles contain neighborhood boundaries for major cities in 41 states plus DC. This fairly new data set comprises nearly 7,000 neighborhoods, but as \citet{schernthanner_spatial_2016} point out, Zillow does not publish the methodology used to construct these boundaries. However, despite its newness it already has a track record in the academic literature: \citet{besbris_effect_2015} use Zillow boundaries to examine neighborhood stigma and \citet{albrecht_indicator_2014} use them to study neighborhood-level poverty in New York.

For each of these geometries, we use OSMnx to download the (drivable, public) street network it contains, a process described in detail in \cite{boeing_osmnx:_2017} and summarized here. First OSMnx buffers each geometry by 0.5 km, then downloads the OpenStreetMap \enquote{nodes} and \enquote{ways} within this buffer. Next it constructs a street network from these data, corrects the topology, calculates street counts per node, then truncates the network to the original, desired polygon (ensuring that intersections are not considered dead-ends simply because an incident edge connects to a node outside the desired polygon). OSMnx saves each of these networks to disk as GraphML and shapefiles. Finally, it calculates metric and topological measures for each network, described below. Such measures extend the toolkit commonly used in urban form studies \cite{talen_measuring_2003, ewing_travel_2010}.

@@

Brief descriptions of these OSMnx-calculated measures are discussed here, but extended technical definitions and algorithms can be found in e.g. \cite{trudeau_introduction_1994, albert_statistical_2002, dorogovtsev_evolution_2002, brandes_network_2005, costa_characterization_2007, newman_structure_2003, newman_networks:_2010, barthelemy_spatial_2011, cranmer_navigating_2017}. \emph{Average street length}, the mean edge length in the undirected representation of the graph, serves as a linear proxy for block size and indicates how fine-grained or coarse-grained the network is. \emph{Node density} is the number of nodes divided by the area covered by the network. \emph{Intersection density} is the density of the set of nodes with more than one street emanating from them (thus excluding dead-ends). \emph{Edge density} is the sum of all edge lengths divided by area, and the physical \emph{street density} is the sum of all edges in the undirected representation of the graph divided by area. These density measures all provide further indication of how fine-grained the network is. Finally, \emph{average circuity} represents the average ratio between an edge length and the straight-line distance between the two nodes it links.

The \emph{average node degree}, or mean number of edges incident to each node, quantifies how well the nodes are connected on average. Similarly, but more concretely, the \emph{average streets per node} measures the mean number of physical streets (i.e., edges in the undirected representation of the graph) that emanate from each intersection and dead-end. This adapts the average node degree for physical form rather than directed flow. The statistical and spatial distributions of number of streets per node characterize the type, prevalence, and dispersion of intersection connectedness in the network. \emph{Connectivity} measures the minimum number of nodes or edges that must be removed from a connected graph to disconnect it. The \emph{average node connectivity} (ANC) of a network---the mean number of internally node-disjoint paths between each pair of nodes in the graph---represents the expected number of nodes that must be removed to disconnect a randomly selected pair of non-adjacent nodes \cite{beineke_average_2002,dankelmann_bounds_2003}.

The \emph{clustering coefficient} of a node is the ratio of the number of edges between its neighbors to the maximum possible number of edges that could exist between these neighbors \cite{jiang_topological_2004, opsahl_clustering_2009}. The weighted clustering coefficient weights this ratio by edge length and the average clustering coefficient is the mean of the clustering coefficients of all the nodes in the network. Centrality indicates the importance of nodes in a network \cite{huang_trajgraph:_2016, zhong_revealing_2017}. \emph{Betweenness centrality} evaluates the number of shortest paths that pass through each node \cite{barthelemy_betweenness_2004}. The maximum betweenness centrality (MBC) in a network specifies the proportion of shortest paths that pass through the most important node. This is an indicator of resilience: networks with a high MBC are more prone to failure or inefficiency should this single choke point fail. Finally, \emph{PageRank} ranks nodes based on the structure of incoming links and the rank of the source node \cite{brin_anatomy_1998, jiang_predicting_2008, agryzkov_algorithm_2012, chin_geographically_2015, gleich_pagerank_2015}.

@@

\begin{figure}
\includegraphics[width=1\textwidth]{media/fig01.png}
\caption{Intersection density per urbanized area (left) and average streets per node per urbanized area (right) from lowest (pale yellow) to highest (dark red), in the contiguous US}
\label{fig01}
\end{figure}

Nevertheless, looking across the data set provides a sense of the breadth of American metropolitan street networks. New York's urbanized area---America's largest-–-has 373,309 intersections and 79 million meters of linear street (or 417,570 and 83.4 million if including service roads). Delano, CA's urbanized area---America's smallest---has 874 intersections and 222,328 meters of linear street (or 964 and 231,000 meters if including service roads). The typical American urbanized area is approximately 185 km\textsuperscript{2} in land area, has 5,830 intersections, and 1.3 million linear meters of street. Its street network is about 7.4\% more circuitous than straight-line as-the-crow-flies edges between nodes would be. The most circuitous network is 14\% more circuitous than straight-line would be, and least is only 2\%. Looking density, grain, and connectivity in the typical urbanized area, the average street segment length (a proxy for block size) is 160 meters. The longest average street segment is the 226-meter average of urbanized Danbury, CT. Puerto Rican cities hold the top four positions for shortest average street segment length, but among the 50 states plus DC, the shortest average street segment is the 125.3-meter average of urbanized Tracy, CA, indicating a much finer street network. The urbanized area of Portland, Oregon, with its famously compact walkable blocks, ranks second at 125.5 meters on average.

\begin{table}

@@

\end{table}

\begin{figure}
\includegraphics[width=0.75\textwidth]{media/fig02.png}
\caption{Distribution of node types in 9 urbanized areas, with number of streets emanating from the node on the $x$-axis and proportion of nodes of this type on the $y$-axis }
\label{fig02}
\end{figure}

@@

The distribution of node types (i.e., intersections and dead ends) provides an indicator of network connectedness. The typical urbanized area averages 2.8 streets per intersection: many 3-way intersections, fewer dead-ends, and even fewer 4-way intersections. The gridlike San Angelo, TX urbanized area has the most streets per node (3.2) on average, and (outside of Puerto Rico, which contains the seven lowest urbanized areas) the sprawling, disconnected Lexington Park, MD urbanized area has the fewest (2.2). These two urbanized areas fit the trend seen in the spatial distribution across the US in Figure \ref{fig01}: urbanized areas in the great plains and Midwest have particularly high numbers of streets per node on average, indicating more gridlike, connected networks. Cities in the southern and western US tend to have fewer streets per node, reflecting more dead-ends and a disconnected network. This finding is discussed in more detail in the upcoming section.

@@

\begin{table}

@@

\end{table}

@@

\begin{table}
\centering
\caption{Selected summary stats for every incorporated city and town in the US.}
\label{table03}
\begin{tabular}{llllll}
	\toprule
	measure                                     & mean    & std dev & min            & median         & max       \\
	\midrule
	Area (km\textsuperscript{2})                                  & 16.703  & 107.499 & 0.039          & 3.918          & 7434.258  \\
	Avg of the avg neighborhood degree          & 2.94    & 0.297   & 0.4            & 2.953          & 3.735     \\
	Avg of the avg weighted neighborhood degree & 0.033   & 0.141   & \textless0.001 & 0.029          & 9.357     \\
	Avg circuity                                & 1.067   & 0.159   & 1              & 1.055          & 20.452    \\
	Avg clustering coefficient                  & 0.048   & 0.041   & \textless0.001 & 0.04           & 1         \\
	Avg weighted clustering coefficient         & 0.01    & 0.018   & \textless0.001 & 0.005          & 0.524     \\
	Intersection count                          & 324     & 1266    & 0              & 83             & 62996     \\
	Avg degree centrality                       & 0.093   & 0.136   & \textless0.001 & 0.052          & 2.667     \\
	Edge density (km/km\textsuperscript{2})                       & 12.654  & 6.705   & 0.006          & 11.814         & 58.603    \\
	Avg edge length (m)                         & 161.184 & 80.769  & 25.822         & 144.447        & 3036.957  \\
	Total edge length (km)                      & 159.067 & 578.521 & 0.052          & 40.986         & 24728.326 \\
	Proportion of dead-ends                     & 0.192   & 0.093   & \textless0.001 & 0.184          & 1         \\
	Proportion of 3-way intersections           & 0.572   & 0.11    & \textless0.001 & 0.579          & 1         \\
	Proportion of 4-way intersections           & 0.237   & 0.129   & \textless0.001 & 0.217          & 1         \\
	Intersection density (per km\textsuperscript{2})              & 29.363  & 21.607  & \textless0.001 & 24.719         & 259.647   \\
	Average node degree                         & 5.251   & 0.668   & 0.8            & 5.268          & 7.166     \\
	m                                           & 1046    & 3924    & 2              & 275            & 176161    \\
	n                                           & 401     & 1516    & 2              & 103            & 71993     \\
	Node density (per km\textsuperscript{2})                      & 35.449  & 24.409  & 0.047          & 30.718         & 296.74    \\
	Max PageRank value                          & 0.034   & 0.046   & \textless0.001 & 0.021          & 0.87      \\
	Min PageRank value                          & 0.005   & 0.018   & \textless0.001 & 0.002          & 0.5       \\
	Self-loop proportion                        & 0.005   & 0.015   & \textless0.001 & \textless0.001 & 1         \\
	Street density (km/km\textsuperscript{2})                     & 6.528   & 3.435   & 0.003          & 6.109          & 29.302    \\
	Average street segment length (m)           & 162.408 & 81.035  & 25.822         & 145.479        & 3036.957  \\
	Total street length (km)                    & 86.096  & 331.048 & 0.026          & 21.005         & 15348.008 \\
	Street segment count                        & 558     & 2208    & 1              & 140            & 107393    \\
	Average streets per node                    & 2.851   & 0.282   & 1              & 2.852          & 4         \\
	\bottomrule
\end{tabular}
\end{table}

>>>>

\documentclass[Afour,sageh,times]{sagej}

\usepackage{graphicx}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{pdflscape}
\usepackage{setspace}
\usepackage[USenglish]{babel}
\usepackage{csquotes}
\usepackage{natbib}
\usepackage{url}

\usepackage{moreverb,url}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
		T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\def\volumeyear{2017}

\renewcommand{\cite}{\citep}
\doublespacing

\begin{document}
\title{A Multi-Scale Analysis of 27,000 Urban Street Networks}
\author{Author Redacted}
\date{March 2017}
\maketitle

@@

OpenStreetMap offers an underexplored source of worldwide geospatial data useful to urban researchers. This study uses the OSMnx software to automatically download and analyze 27,009 US street networks from OpenStreetMap at metropolitan, municipal, and neighborhood scales---namely, every US city and town ($N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$). It presents wide-ranging empirical findings on US urban form and street network characteristics, emphasizing measures relevant to graph theory, urban design, and morphology such as structure, connectedness, density, centrality, and resilience. In the past, street network data acquisition and processing has been challenging and ad hoc. This study illustrates the use of OSMnx and OpenStreetMap to consistently conduct street network analysis with extremely large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. These street networks and measures data sets have been shared in a public repository for other researchers to use.

@@

On May 20, 1862, Abraham Lincoln signed the Homestead Act into law, making land across the United States' Midwest and Great Plains available for free to applicants \cite{porterfield_homestead_2005}. Under its auspices over the next 70 years, the federal government distributed 10\% of the entire US landmass to private owners in the form of 1.6 million homesteads \cite{lee_kansas_1979, sherraden_inclusion_2005}. New towns with gridiron street networks sprang up rapidly across the Great Plains and Midwest, due to both the prevailing urban design paradigm of the day and the standardized rectilinear town plats used repeatedly to lay out instant new cities \cite{southworth_streets_1997}. Through path dependence, the spatial signatures of these land use laws, design paradigms, and planning instruments can still be seen today in these cities' urban forms and street networks. Cross-sectional analysis of American urban form can reveal these artifacts and histories through street networks at metropolitan, municipal, and neighborhood scales.

Network analysis is a natural approach to the study of cities as complex systems \cite{masucci_random_2009}. The empirical literature on street networks is growing ever richer, but suffers from some limitations---discussed in detail in \cite{redacted_osmnx:_2017} and summarized here. First, sample sizes tend to be fairly small due to data availability, gathering, and processing constraints: most studies in this literature that conduct topological or metric analyses tend to have sample sizes ranging around 10 to 50 networks \cite{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, strano_urban_2013}, which may limit the generalizability and interpretability of findings. Second, reproducibility has been difficult when the dozens of decisions that go into analysis---such as spatial extents, topological simplification and correction, definitions of nodes and edges, etc.---are ad hoc or only partly reported \citep[e.g.,][]{porta_network_2006, strano_urban_2013}. Third, and related to the first two, studies frequently oversimplify to planar or undirected primal graphs for tractability \cite[e.g.,][]{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, masucci_random_2009}, or use dual graphs despite the loss of geographic, metric information \cite{batty_network_2005, jiang_integration_2002, ratti_space_2004, crucitti_centrality_2006, crucitti_centrality_2006-1}.

This study addresses these limitations by conducting a morphological analysis of urban street networks at multiple scales, with large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. In particular, it examines 27,009 urban street networks---represented as primal, nonplanar, weighted multidigraphs with possible self-loops---at multiple overlapping scales across the US, focusing on structure, connectedness, centrality, and resilience. It examines the street networks of every incorporated city and town (population size $N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$) in the US. To do so, it uses OSMnx---a new street network research toolkit \cite{redacted_osmnx:_2017}---to download, construct, and analyze these street networks at metropolitan, municipal, and neighborhood scales. These street networks and measures data sets have been compiled and shared in a public repository for other researchers to repurpose at [redacted].

The purpose of this paper is threefold. First, it describes and demonstrates a new methodology for easily and consistently acquiring, constructing, and analyzing large samples of street networks as nonplanar directed graphs. Second, it presents empirical findings of descriptive urban morphology for the street networks of every US city, urbanized area, and Zillow neighborhood. Third, it investigates with large sample sizes some claims previously made in the research literature using small sample sizes. This paper is organized as follows. In the next section, it discusses the data sources, tools, and methods used to collect, construct, and analyze these street networks. Then it presents findings of the analyses at metropolitan, municipal, and neighborhood scales. Finally, it concludes with a discussion of these findings and their implications for street network analysis, urban morphology, and city planning.

@@

A network (also called a \emph{graph} in mathematics) comprises a set of nodes connected to one another by a set of edges. Street networks can be conceptualized as primal, directed, nonplanar graphs. A \emph{primal} street network represents intersections as nodes and street segments as edges. A \emph{directed} network has directed edges: that is, edge $uv$ points one-way from node $u$ to node $v$, but there need not exist a reciprocal edge $vu$. A \emph{planar} network can be represented in two dimensions with its edges intersecting only at nodes \cite{viana_simplicity_2013, fischer_spatial_2014}. Most street networks are nonplanar---due to grade-separated expressways, overpasses, bridges, tunnels, etc.---but most quantitative studies of urban street networks represent them as planar \cite[e.g.,][]{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, masucci_random_2009, strano_urban_2013} for tractability because bridges and tunnels are uncommon in some cities. Planar graphs may reasonably model the street networks of old European town centers, but poorly model the street networks of modern autocentric cities like Los Angeles or Tokyo with many grade-separated expressways, bridges, and underpasses.

@@

This study uses OSMnx to download, construct, correct, analyze, and visualize street networks at metropolitan, municipal, and neighborhood scales. OSMnx is a Python-based research tool that easily downloads OpenStreetMap data for any place name, address, or polygon in the world, then constructs it into a spatially-embedded graph-theoretic object for analysis and visualization \cite{redacted_osmnx:_2017}. OpenStreetMap is a collaborative worldwide mapping project that makes its spatial data available via various APIs \cite{corcoran_analysing_2013, jokar_arsanjani_openstreetmap_2015}. These data are of high quality and compare favorably to CIA World Factbook estimates and US Census TIGER/Line data \cite{haklay_how_2010, over_generating_2010, zielstra_comparative_2011, maron_how_2015, wu_improving_2005, frizzelle_importance_2009}. In 2007, OpenStreetMap imported the TIGER/Line roads (2005 vintage) and since then, many community-led corrections and improvements have been made \cite{willis_openstreetmap_2008}. Many of these additions go beyond TIGER/Line's scope, including passageways between buildings, footpaths through parks, bike routes, and detailed feature attributes such as finer-grained street classifiers, speed limits, etc.

To define the study sites and their spatial boundaries, we use three sets of geometries. The first defines the metropolitan-scale study sites using the 2016 TIGER/Line shapefile of US census bureau urban areas. Each census-defined urban area comprises a set of tracts that meet a minimum density threshold \cite{u.s._census_bureau_2010_2010}. We retain only the \emph{urbanized areas} subset of these data (i.e., areas with greater than 50,000 population), discarding the small \emph{urban clusters} subset. The second set of geometries defines our municipal-scale study sites using 51 separate TIGER/Line shapefiles (again, 2016) of US census bureau \emph{places} within all 50 states plus DC. We discard the subset of \emph{census-designated places} (i.e., small unincorporated communities) in these data, while retaining every US city and town. The third set of geometries defines the neighborhood-scale study sites using 42 separate shapefiles from Zillow, a real estate database company. These shapefiles contain neighborhood boundaries for major cities in 41 states plus DC. This fairly new data set comprises nearly 7,000 neighborhoods, but as \citet{schernthanner_spatial_2016} point out, Zillow does not publish the methodology used to construct these boundaries. However, despite its newness it already has a track record in the academic literature: \citet{besbris_effect_2015} use Zillow boundaries to examine neighborhood stigma and \citet{albrecht_indicator_2014} use them to study neighborhood-level poverty in New York.

For each of these geometries, we use OSMnx to download the (drivable, public) street network within it, a process described in detail in \cite{redacted_osmnx:_2017} and summarized here. First OSMnx buffers each geometry by 0.5 km, then downloads the OpenStreetMap \enquote{nodes} and \enquote{ways} within this buffer. Next it constructs a street network from these data, corrects the topology, calculates street counts per node, then truncates the network to the original, desired polygon. OSMnx saves each of these networks to disk as GraphML and shapefiles. Finally, it calculates metric and topological measures for each network, summarized below. Such measures extend the toolkit commonly used in urban form studies \cite{talen_measuring_2003, ewing_travel_2010}.

@@

Brief descriptions of these OSMnx-calculated measures are discussed here, but extended technical definitions and algorithms can be found in e.g. \cite{trudeau_introduction_1994, albert_statistical_2002, dorogovtsev_evolution_2002, brandes_network_2005, costa_characterization_2007, newman_structure_2003, newman_networks:_2010, barthelemy_spatial_2011, cranmer_navigating_2017}. \emph{Average street length}, the mean edge length in the undirected representation of the graph, serves as a linear proxy for block size and indicates how fine-grained or coarse-grained the network is. \emph{Node density} is the number of nodes divided by the area covered by the network. \emph{Intersection density} is the density of the set of nodes with more than one street emanating from them (thus excluding dead-ends). \emph{Edge density} is the sum of all edge lengths divided by area, and the physical \emph{street density} is the sum of all edges in the undirected representation of the graph divided by area. These density measures suggest how fine-grained the network is. Finally, \emph{average circuity} represents the average ratio between each edge length and the straight-line distance between the two nodes it links.

The \emph{average node degree}, or mean number of edges incident to each node, quantifies how well the nodes are connected. Similarly, but more concretely, the \emph{average streets per node} measures the mean number of physical streets (i.e., edges in the undirected representation of the graph) that emanate from each intersection and dead-end. This adapts the average node degree for physical form rather than directed flow. The statistical and spatial distributions of number of streets per node characterize the type, prevalence, and dispersion of intersection connectedness in the network. \emph{Connectivity} measures the minimum number of nodes or edges that must be removed from a connected network to disconnect it. The \emph{average node connectivity} (ANC) of a network---the mean number of internally node-disjoint paths between each pair of nodes---represents the expected number of nodes that must be removed to disconnect a randomly selected pair of non-adjacent nodes \cite{beineke_average_2002,dankelmann_bounds_2003}.

The \emph{clustering coefficient} of a node is the ratio of the number of edges between its neighbors to the maximum possible number of edges that could exist between these neighbors \cite{jiang_topological_2004, opsahl_clustering_2009}. The weighted clustering coefficient weights this ratio by edge length and the average clustering coefficient is the mean of the clustering coefficients of all the network's nodes. \emph{Betweenness centrality} indicates the importance of nodes in a network by evaluating the number of shortest paths that pass through each node \cite{barthelemy_betweenness_2004,huang_trajgraph:_2016, zhong_revealing_2017}. The maximum betweenness centrality (MBC) in a network specifies the proportion of shortest paths that pass through the most important node. This is an indicator of resilience: networks with a high MBC are more prone to failure or inefficiency should this single choke point fail. Finally, \emph{PageRank} ranks nodes based on the structure of incoming links and the rank of the source node \cite{brin_anatomy_1998, jiang_predicting_2008, agryzkov_algorithm_2012, chin_geographically_2015, gleich_pagerank_2015}.

@@

Nevertheless, looking across the data set provides a sense of the breadth of American metropolitan street networks. New York's urbanized area---America's largest-–-has 373,309 intersections and 79 million meters of linear street (or 417,570 and 83.4 million if including service roads). Delano, CA's urbanized area---America's smallest---has 874 intersections and 222,328 meters of linear street (or 964 and 231,000 meters if including service roads). The typical American urbanized area is approximately 185 km\textsuperscript{2} in land area, has 5,830 intersections, and 1.3 million linear meters of street. Its street network is about 7.4\% more circuitous than straight-line as-the-crow-flies edges between nodes would be. The most circuitous network is 14\% more circuitous than straight-line would be, and the least is only 2\%. Looking at density, grain, and connectivity in the typical urbanized area, the average street segment length (a proxy for block size) is 160 meters. The longest average street segment is the 226-meter average of urbanized Danbury, CT. Puerto Rican cities hold the top four positions for shortest average street segment length, but among the 50 states plus DC, the shortest average street segment is the 125.3-meter average of urbanized Tracy, CA, indicating a fine-grained street network. The urbanized area of Portland, Oregon, with its famously compact walkable blocks, ranks second at 125.5 meters.

\begin{table*}

@@

\end{table*}

@@

\begin{figure*}
	\includegraphics[width=1\textwidth]{fig01.png}
	\caption{Intersection density per urbanized area (left) and average streets per node per urbanized area (right) from lowest (pale yellow) to highest (dark red), in the contiguous US}
	\label{fig01}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.7\textwidth]{fig02.png}
	\caption{Distribution of node types in 9 urbanized areas, with number of streets emanating from the node on the $x$-axis and proportion of nodes of this type on the $y$-axis }
	\label{fig02}
\end{figure*}

The distribution of node types (i.e., intersections and dead ends) provides an indicator of network connectedness. The typical urbanized area averages 2.8 streets per intersection: many 3-way intersections, fewer dead-ends, and even fewer 4-way intersections. The gridlike San Angelo, TX urbanized area has the most streets per node (3.2) on average, and (outside of Puerto Rico, which contains the seven lowest urbanized areas) the sprawling, disconnected Lexington Park, MD urbanized area has the fewest (2.2). These fit the trend seen in the spatial distribution across the US in Figure \ref{fig01}: urbanized areas in the Great Plains and Midwest have particularly high numbers of streets per node on average, indicating more gridlike, connected networks. Cities in the southern and western US tend to have fewer streets per node, reflecting more dead-ends and a disconnected network. This finding is discussed in more detail in the upcoming section.

@@

\begin{table*}

@@

\end{table*}

@@

\begin{table*}
\centering
\caption{Selected summary stats for every incorporated city and town in the US.}
\label{table03}
\begin{tabular}{llllll}
	\toprule
	measure                                     & mean    & std dev & min            & median         & max       \\
	\midrule
	Area (km\textsuperscript{2})                                  & 16.703  & 107.499 & 0.039          & 3.918          & 7434.258  \\
	Avg of the avg neighborhood degree          & 2.94    & 0.297   & 0.4            & 2.953          & 3.735     \\
	Avg of the avg weighted neighborhood degree & 0.033   & 0.141   & \textless0.001 & 0.029          & 9.357     \\
	Avg circuity                                & 1.067   & 0.159   & 1              & 1.055          & 20.452    \\
	Avg clustering coefficient                  & 0.048   & 0.041   & \textless0.001 & 0.04           & 1         \\
	Avg weighted clustering coefficient         & 0.01    & 0.018   & \textless0.001 & 0.005          & 0.524     \\
	Intersection count                          & 324     & 1266    & 0              & 83             & 62996     \\
	Avg degree centrality                       & 0.093   & 0.136   & \textless0.001 & 0.052          & 2.667     \\
	Edge density (km/km\textsuperscript{2})                       & 12.654  & 6.705   & 0.006          & 11.814         & 58.603    \\
	Avg edge length (m)                         & 161.184 & 80.769  & 25.822         & 144.447        & 3036.957  \\
	Total edge length (km)                      & 159.067 & 578.521 & 0.052          & 40.986         & 24728.326 \\
	Proportion of dead-ends                     & 0.192   & 0.093   & \textless0.001 & 0.184          & 1         \\
	Proportion of 3-way intersections           & 0.572   & 0.11    & \textless0.001 & 0.579          & 1         \\
	Proportion of 4-way intersections           & 0.237   & 0.129   & \textless0.001 & 0.217          & 1         \\
	Intersection density (per km\textsuperscript{2})              & 29.363  & 21.607  & \textless0.001 & 24.719         & 259.647   \\
	Average node degree                         & 5.251   & 0.668   & 0.8            & 5.268          & 7.166     \\
	m                                           & 1046    & 3924    & 2              & 275            & 176161    \\
	n                                           & 401     & 1516    & 2              & 103            & 71993     \\
	Node density (per km\textsuperscript{2})                      & 35.449  & 24.409  & 0.047          & 30.718         & 296.74    \\
	Max PageRank value                          & 0.034   & 0.046   & \textless0.001 & 0.021          & 0.87      \\
	Min PageRank value                          & 0.005   & 0.018   & \textless0.001 & 0.002          & 0.5       \\
	Self-loop proportion                        & 0.005   & 0.015   & \textless0.001 & \textless0.001 & 1         \\
	Street density (km/km\textsuperscript{2})                     & 6.528   & 3.435   & 0.003          & 6.109          & 29.302    \\
	Average street segment length (m)           & 162.408 & 81.035  & 25.822         & 145.479        & 3036.957  \\
	Total street length (km)                    & 86.096  & 331.048 & 0.026          & 21.005         & 15348.008 \\
	Street segment count                        & 558     & 2208    & 1              & 140            & 107393    \\
	Average streets per node                    & 2.851   & 0.282   & 1              & 2.852          & 4         \\
	\bottomrule
\end{tabular}
\end{table*}
